Tidal streams in triaxial systems

### 2015-02-20 ms.tex
lyapunov

\subsection{Potential choice}

@@

Computing the Lyapunov spectrum for a given orbit is often not necessary if one is only interested in characterizing the degree of chaos. Instead, it is often more efficient to compute the maximum Lyapunov exponent by estimating the finite-time maximum Lyapunov exponent (FTMLE), defined as 
\begin{equation}
	l_{\rm max}(t) = \frac{1}{t}\ln \frac{\|\delta \bs{w}(t)\|}{\|\delta \bs{w}_0\|}.\label{eq:lmax}
\end{equation}
The maximum Lyapunov exponent is the limit as $t\rightarrow \infty$ of the FTMLE
\begin{equation}
	\lambda_{\rm max} = \lim_{t\rightarrow\infty}l_{\rm max}(t). \label{eq:lyapmax}
\end{equation}
Numerically computing this quantity is not trivial because (1) obviously the limit to infinity is not possible and (2) the norm of the deviation vector $\|\delta \bs{w}(t)\|$ is expected to increase exponentially for chaotic orbits, leading to numerical problems. To circumvent these issues, it is sufficient to instead start a nearby orbit with some small initial deviation with norm $\delta_0$, integrate for a sufficiently small amount of time, $\tau$, then renormalize the deviation back to the initial norm (e.g., cite Bennetin et al. 1976, Tabor 1989). 
%The following pseudocode outlines this procedure:\footnote{see Gary for Python implementation?}\\
%\begin{algorithmic}[1]
%\State {\bf define} orbital integration timestep, $h$, and number of steps, $K$
%\State {\bf define} initial norm of deviation vector, $\delta_0$, to be sufficiently small
%\State {\bf define} renormalization integration period, $\tau \ll hK$ 
%\For{each timestep when integrating the main orbit, $\bs{w}(t)$}
%\State step forward the orbit and deviation vector orbit by one timestep, $t_{i-1} \rightarrow t_i$
%\If {a normalization timestep}
%\State measure and store the length of the deviation vector, $\delta_i = \|\delta \bs{w}_i\|$
%\State renormalize the length of the deviation vector, $\delta \bs{w}_i = \delta \bs{w}_i (\delta_0/\delta_i)$
%\EndIf
%\EndFor 
%\end{algorithmic}
The FTMLE after a given number of timesteps, $N$, is then estimated as
\begin{equation}
	l_N = \frac{1}{N\tau}\sum_i^N \ln \, \delta_i
\end{equation}
and the MLE is
\begin{equation}
	\lambda_{\rm max} = \lim_{N\rightarrow \infty} l_N.
\end{equation}
[Some more practical issues with implementation] [Figure showing $l_N$ for regular, chaotic orbit?]

[Copy in text fomr ``behavior of MLE'']

[Computationally, expensive to compute Lyapunov exponent for sticky or mildly chaotic orbits]

>>>>

\subsection{Potential choice}\label{sec:potential}

@@

Computing the Lyapunov spectrum for a given orbit is often not necessary if one is only interested in characterizing the degree of chaos. Instead, it is often more efficient to compute the maximum Lyapunov exponent by estimating the finite-time maximum Lyapunov exponent (FTMLE), defined as 
\begin{equation}
	l_{\rm max}(t) = \frac{1}{t}\ln \frac{\|\delta \bs{w}(t)\|}{\|\delta \bs{w}_0\|}.\label{eq:lmax}
\end{equation}
The maximum Lyapunov exponent is the limit as $t\rightarrow \infty$ of the FTMLE
\begin{equation}
	\lambda_{\rm max} = \lim_{t\rightarrow\infty}l_{\rm max}(t). \label{eq:lyapmax}
\end{equation}
Numerically computing this quantity is not trivial because (1) obviously the limit to infinity is not possible and (2) the norm of the deviation vector $\|\delta \bs{w}(t)\|$ is expected to increase exponentially for chaotic orbits, leading to numerical problems. To circumvent these issues, it is sufficient to instead start a nearby orbit with some small initial deviation with norm $\delta_0$, integrate for a sufficiently small amount of time, $\tau$, then renormalize the deviation back to the initial norm (e.g., cite Bennetin et al. 1976, Tabor 1989). 
%The following pseudocode outlines this procedure:\footnote{see Gary for Python implementation?}\\
%\begin{algorithmic}[1]
%\State {\bf define} orbital integration timestep, $h$, and number of steps, $K$
%\State {\bf define} initial norm of deviation vector, $\delta_0$, to be sufficiently small
%\State {\bf define} renormalization integration period, $\tau \ll hK$ 
%\For{each timestep when integrating the main orbit, $\bs{w}(t)$}
%\State step forward the orbit and deviation vector orbit by one timestep, $t_{i-1} \rightarrow t_i$
%\If {a normalization timestep}
%\State measure and store the length of the deviation vector, $\delta_i = \|\delta \bs{w}_i\|$
%\State renormalize the length of the deviation vector, $\delta \bs{w}_i = \delta \bs{w}_i (\delta_0/\delta_i)$
%\EndIf
%\EndFor 
%\end{algorithmic}
The FTMLE after a given number of timesteps, $N$, is then estimated as
\begin{equation}
	l_N = \frac{1}{N\tau}\sum_i^N \ln \, \delta_i
\end{equation}
and the MLE is
\begin{equation}
	\lambda_{\rm max} = \lim_{N\rightarrow \infty} l_N.
\end{equation}

For most regular orbits, deviations will grow close to linearly or as some power-law of time. 
%As we have seen (Section~\ref{sec:introduction}), if the orbit is regular, there exists a local transformation to action-angle variables where the angle variables increase linearly with time, $\theta_n \propto \Omega_n t$. We can look at small variations around the orbit defined by the above equations:\footnote{Note the implied summation in repeated indices in the above expression.}
%\begin{align}
%	\frac{d (\delta \theta_n)}{dt} &= \frac{\partial \Omega_n}{\partial J_m} \delta J_m
%\end{align}
%These equations are trivially integrated:
%\begin{align}
%	\delta \theta_n(t) &= \delta \theta_n(0) + \left(\frac{\partial \Omega_n}{\partial J_m} \delta J_m \right) t
%\end{align}
%It's clear then that the norm of the deviation vector grows linearly with time:
%\begin{align}
%	\|\delta \bs{w}(t)\| &= \left[\sum_n (\delta \theta_n(t))^2 + \sum_n (\delta J_n)^2\right]^{1/2}\\
%	&= \left[\sum_n \left(\delta \theta_n(0) + \frac{\partial \Omega_n}{\partial J_m} \delta J_m t\right)^2 + \sum_n (\delta J_n)^2\right]^{1/2}\\
%	&\propto t.
%\end{align}
From Eq.~\ref{eq:lmax} and Eq.~\ref{eq:lyapmax} it is evident that any deviation vector that grows as a power law with time, $t^k$, will asymptote to 0 from the limit 
\begin{equation}
	\lambda_{\rm max} \propto \lim_{t\rightarrow \infty} k \frac{\ln t}{t} = 0.
\end{equation}
At long times, the implication is that the the numerically computed MLE should approach 0 like $t^{-1}$ for regular orbits. For chaotic orbits, the divergence is exponential, and the limit should converge to the rate of the exponential: the Lyapunov exponent.

In practice, the MLE is often estimated as the mean of $l_N$ after the summation diverges from power-law behavior. Figure~\ref{fig:lyap-orbits} shows $l_N$ computed for two nearby orbits in the potential of Section~\ref{sec:potential}. In physical units and Cartesian coordinates, the initial conditions for the two orbits are:
\begin{align}
	\bs{x}_{0}^{(1)} &= (27.85, 0, 23.35)~{\rm kpc}\\
	\bs{v}_{0}^{(1)} &= (0, 115.37, 0)~{\rm km~s}^{-1}\\
	\bs{x}_{0}^{(2)} &= (27.85, 0, 23.10)~{\rm kpc}\\
	\bs{v}_{0}^{(2)} &= (0, 116.43, 0)~{\rm km~s}^{-1}.
\end{align}
Over-plotted on both panels are lines (dashed) that show $t^{-0.9}$ behavior. The MLE estimate continues towards zero for the regular orbit (top) but saturates for the chaotic orbit (bottom) and continues to increase at least until the end of our integration period. For mildly chaotic or ``sticky'' orbits, it can take integration periods of many hundreds or thousands of orbits to determine a reliable estimate of the Lyapunov exponent.

\begin{figure*}[!h]
\begin{center}
\includegraphics[width=0.75\textwidth]{figures/lyap-orbits.pdf}
\caption{ } \label{fig:lyap-orbits}
\end{center}
\end{figure*}
